---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

To begin, we load the provided data into the `activity` data frame.

```{r echo=TRUE}
data_filename <- unzip("activity.zip", list = TRUE)[[1]] 
unzip("activity.zip")
activity <- read.csv(
    file = data_filename, 
    colClasses = c("integer", "Date", "integer")
)
```

The interval data is massaged to a time column (in "HH:MM" format) in the data frame. 

```{r}
library(stringr)
activity$time <- str_pad(activity$interval, width=4, side = "left", pad = "0")
activity$time <- paste(
    substr(activity$time, 1, 2),
    substr(activity$time, 3, 4),
    sep = ":"
)
head(activity)
```

## What is mean total number of steps taken per day?

We create a histogram of the **total** steps taken per day.

```{r}
total.steps.by.date <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
hist(total.steps.by.date, breaks = 20) # custom breaks for increased detail
```

As can be seen below, the **mean** and **median** total steps taken per day are 9354 and 10400, respectively.

```{r}
summary(total.steps.by.date, na.rm = TRUE)
```

## What is the average daily activity pattern?

The time-series plot below illustrates the average number of steps taken at each of the 5-minute intervals.

```{r}
# calculate average steps by 5-minute interval
average.steps.by.interval <- tapply(
    activity$steps,
    activity$time,
    mean,
    na.rm = TRUE
)

# plot calculated average steps
plot(
    average.steps.by.interval,
    type = "l", 
    main = "Average Steps Taken by Five-Minute Interval",
    xlab = "Time Interval",
    xaxt = "n",
    ylab = "Steps"
)

# change x-axis labels to reflect values, not indices
xnames <- names(average.steps.by.interval)
axis(1, at=1:length(xnames), labels=xnames)
```

The **`r names(which.max(average.steps.by.interval))`** interval has the highest average number of steps taken across all days in the data set (viz., `r signif(max(average.steps.by.interval, na.rm = TRUE), digits = 1)`).

## Inputting missing values



## Are there differences in activity patterns between weekdays and weekends?
